---
title: "BlueCarbon Compaction Correction — Method Comparison"
subtitle: "BlueCarbon package vs VM0033 raw-depth pipeline"
date: today
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
if (!file.exists("_targets_bluecarbon")) setwd("..")
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bc_decompacted  <- tar_read(bc_decompacted,               store = "_targets_bluecarbon")
bc_stocks       <- tar_read(bc_stocks,                    store = "_targets_bluecarbon")
bc_extrap       <- tar_read(bc_extrapolation_test,        store = "_targets_bluecarbon")
bc_comparison   <- tar_read(bc_vm_comparison,             store = "_targets_bluecarbon")
bc_comp_plot    <- tar_read(bc_comparison_plot,           store = "_targets_bluecarbon")
bc_max_depth    <- tar_read(bc_max_depth,                 store = "_targets_bluecarbon")

vm_summary <- tryCatch(
  tar_read(stratum_summary, store = "_targets"),
  error = function(e) NULL
)

has_vm <- !is.null(vm_summary) && !is.null(bc_comparison)
```

## What this report shows

Percussion coring compresses sediment during collection. Without correction,
depth-referenced carbon stocks will be underestimated because compressed samples
are assigned shallower depths than they occupied in situ.

This report applies **BlueCarbon package** compaction correction
(`BlueCarbon::decompact()`) to the field cores and compares the resulting
carbon stocks against the **VM0033 raw-depth pipeline** (`tar_make()`).

The comparison answers: *how much does compaction correction change the
estimated carbon stock at this site?*

---

## Compaction correction

### Compaction per core

```{r}
comp_tbl <- bc_decompacted |>
  distinct(core_id, compaction) |>
  arrange(core_id) |>
  mutate(
    compaction_pct = round(compaction, 1),
    corrected      = compaction_pct > 0
  )

comp_tbl |>
  select(
    `Core`               = core_id,
    `Compaction (%)`     = compaction_pct,
    `Correction applied` = corrected
  ) |>
  kable(caption = "Estimated compaction per core from field coring measurements.")
```

```{r fig.cap="Compaction percentage by core. Higher values indicate greater compression during collection."}
ggplot(comp_tbl, aes(x = reorder(core_id, -compaction_pct), y = compaction_pct,
                     fill = compaction_pct > 0)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = "#cccccc", "TRUE" = "#2c7a4b")) +
  labs(x = NULL, y = "Compaction (%)") +
  theme_minimal(base_size = 12)
```

### Effect of decompaction on sample depths

Compaction correction shifts sample depths upward (decompacted sediment
occupies more space). The plot below shows original vs corrected midpoint
depths for each sample.

```{r fig.cap="Original vs decompacted sample midpoint depth per core. Points below the 1:1 line indicate upward depth correction."}
depth_compare <- bc_decompacted |>
  mutate(
    depth_mid_orig = (depth_top_cm   + depth_bottom_cm)   / 2,
    depth_mid_corr = (mind_corrected + maxd_corrected)     / 2
  )

ggplot(depth_compare, aes(x = depth_mid_orig, y = depth_mid_corr, colour = core_id)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.7, size = 1.8) +
  labs(
    x      = "Original depth midpoint (cm)",
    y      = "Decompacted depth midpoint (cm)",
    colour = "Core"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

---

## OC stocks per core

Total OC stock estimated by BlueCarbon linear extrapolation to
**`r bc_max_depth` cm** (the deepest core in this dataset). Columns:

- **Stock to depth** — accumulated OC in the top `r bc_max_depth` cm
- **Whole-core stock** — accumulated OC over the entire measured core length
- **SE** — standard error of the linear extrapolation

```{r}
bc_stocks |>
  arrange(stratum, core_id) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  select(
    Stratum                  = stratum,
    Core                     = core_id,
    `Stock to depth (kg/m²)` = stock_kg_m2,
    `Whole-core (kg/m²)`     = stockwc_kg_m2,
    `SE (kg/m²)`             = stock_se_kg_m2
  ) |>
  kable(caption = paste0("Per-core OC stocks to ", bc_max_depth, " cm depth."))
```

```{r fig.cap="Per-core OC stocks. Bar = stock to target depth; point = whole-core stock."}
ggplot(bc_stocks, aes(x = reorder(core_id, stock_kg_m2), y = stock_kg_m2,
                      fill = stratum)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_point(aes(y = stockwc_kg_m2), shape = 21, fill = "white",
             colour = "grey30", size = 3) +
  geom_errorbar(aes(ymin = stock_kg_m2 - stock_se_kg_m2,
                    ymax = stock_kg_m2 + stock_se_kg_m2),
                width = 0.2, colour = "grey40") +
  scale_fill_manual(values = c(IM = "#2c7a4b", NM = "#6baed6", MF = "#fd8d3c"),
                    na.value = "#aaaaaa") +
  coord_flip() +
  labs(x = NULL, y = "OC stock (kg C/m²)", fill = "Stratum") +
  theme_minimal(base_size = 12)
```

---

## Comparison with VM0033 pipeline

```{r}
#| results: asis
if (!has_vm) {
  cat("> **VM0033 pipeline store not found.**  \n",
      "> Run `targets::tar_make()` first, then re-render this report to see the method comparison.\n")
}
```

```{r eval=has_vm, fig.cap="Mean OC stock per stratum: BlueCarbon (decompacted) vs VM0033 (raw depths). Percentage label shows the relative difference attributable to compaction correction."}
if (has_vm) bc_comp_plot
```

```{r eval=has_vm}
if (has_vm) {
  bc_comparison |>
    mutate(across(where(is.numeric), \(x) round(x, 3))) |>
    select(
      Stratum                      = stratum,
      `N cores`                    = n_cores,
      `BlueCarbon mean (kg/m²)`    = mean_bc_kg_m2,
      `BlueCarbon SD (kg/m²)`      = sd_bc_kg_m2,
      `VM0033 mean (kg/m²)`        = mean_vm0033_kg_m2,
      `Difference (kg/m²)`         = diff_kg_m2,
      `Difference (%)`             = diff_pct
    ) |>
    kable(caption = "Per-stratum mean OC stock: BlueCarbon vs VM0033.")
}
```

### Interpreting the difference

A **positive difference** (BlueCarbon > VM0033) indicates that compaction
correction increases stock estimates — compressed sediment had higher bulk
density per unit compressed depth, but once expanded the per-depth stock is
larger. A **negative difference** is unusual and may indicate that compaction
was minimal or that the linear extrapolation underestimated stocks.

---

## Extrapolation validation

BlueCarbon uses a linear regression of cumulative OC vs depth to extrapolate
short cores to the target depth (`r bc_max_depth` cm). This section quantifies
how much error that extrapolation introduces.

```{r fig.cap="Extrapolation test: error (%) introduced by linear extrapolation at varying core truncation fractions. Lower error = better extrapolation quality."}
# bc_extrap may be a ggplot directly or a list with $plot
if (inherits(bc_extrap, "gg")) {
  bc_extrap
} else if (is.list(bc_extrap) && "plot" %in% names(bc_extrap)) {
  bc_extrap$plot
} else {
  ggplot() +
    labs(title = "Extrapolation test not available",
         subtitle = paste0("Requires at least one core reaching ", bc_max_depth, " cm."),
         x = NULL, y = NULL) +
    theme_minimal()
}
```

---

## Using decompacted data in the spatial pipeline

To run the full VM0033 spatial pipeline on compaction-corrected cores:

```r
# Copy the processed file to the data_raw folder
file.copy(
  "outputs/bluecarbon/core_samples_bc_processed.csv",
  "Pre-Analysis Data Preparation/data_raw/core_samples.csv",
  overwrite = TRUE
)

# Invalidate and rebuild all downstream targets
targets::tar_make()
```

---

*Rendered: `r format(Sys.time(), "%Y-%m-%d %H:%M")` |
BlueCarbon package: `r packageVersion("BlueCarbon")`*
