---
title: "Carbon Sequestration Rates"
subtitle: "Radiometric age-depth chronology and OC accumulation rates"
date: today
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
if (!file.exists("_targets_seqrates")) setwd("..")
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

sr_cores   <- tar_read(sr_cores_with_ages, store = "_targets_seqrates")
sr_rates   <- tar_read(sr_rates,           store = "_targets_seqrates")
sr_summary <- tar_read(sr_summary,         store = "_targets_seqrates")
sr_plot    <- tar_read(sr_plot,            store = "_targets_seqrates")

has_rates <- !is.null(sr_rates) && nrow(sr_rates) > 0
```

## What this report shows

Carbon sequestration rate is the mean rate at which organic carbon has
accumulated in the soil column over a defined time window, estimated from
radiometric age-depth chronology data (^210^Pb or ^14^C dated horizons).

**Method:**

1. Sample depths are decompaction-corrected (same as the BlueCarbon pipeline).
2. Ages at each sample midpoint are assigned by **linear interpolation** between
   chronology anchor points; samples outside the dated range are extrapolated
   using the nearest-end slope.
3. BlueCarbon's `estimate_seq_rate()` integrates OC accumulation over the
   specified timeframe (default **`r if(has_rates) sr_rates$timeframe_yr[1] else 100` years**).

> **Note:** Linear interpolation is a simplified age model. For Bayesian
> age-depth modelling with uncertainty, use Bacon or Bchron externally
> and import the output ages in the same `core_chronology.csv` format.

---

## Chronology overview

### Cores with chronology data

```{r}
chrono_summary <- sr_cores |>
  filter(!is.na(age)) |>
  group_by(core_id) |>
  summarise(
    n_samples     = n(),
    max_depth_cm  = round(max(depth_mid, na.rm = TRUE), 1),
    max_age_yr    = round(max(age,       na.rm = TRUE), 0),
    mean_acc_rate = round(max(age, na.rm=TRUE) / max(depth_mid, na.rm=TRUE), 1),
    .groups = "drop"
  )

chrono_summary |>
  select(
    Core                             = core_id,
    `Samples with ages`              = n_samples,
    `Max depth (cm)`                 = max_depth_cm,
    `Max age (yr BP)`                = max_age_yr,
    `Mean accumulation (yr/cm)`      = mean_acc_rate
  ) |>
  kable(caption = "Summary of cores with radiometric chronology data.")
```

### Age-depth relationships

```{r fig.height=4, fig.cap="Age-depth profile for each chronology core. Lines show the interpolated age model used to assign ages to samples."}
sr_cores |>
  filter(!is.na(age)) |>
  ggplot(aes(x = age, y = depth_mid, colour = core_id, group = core_id)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_y_reverse(name = "Depth (cm)") +
  scale_x_continuous(name = "Age (years BP)") +
  scale_colour_brewer(palette = "Set2", name = "Core") +
  labs(title = "Age-depth model (linear interpolation)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

---

## Sequestration rates

```{r results="asis"}
if (!has_rates) {
  cat("> **No sequestration rates could be calculated.**  \n",
      "> This occurs when no core spans the full timeframe ",
      if (!is.null(sr_rates)) sr_rates$timeframe_yr[1] else 100,
      " years.  \n",
      "> Add deeper cores or increase age-depth coverage to enable rate estimation.\n")
}
```

### Per-core rates

```{r eval=has_rates, fig.cap="Carbon sequestration rates per core. Bar = rate over the target timeframe; circle = whole-core rate (total OC / max core age)."}
if (has_rates) sr_plot
```

```{r eval=has_rates}
if (has_rates) {
  sr_rates |>
    mutate(across(where(is.numeric), \(x) round(x, 1))) |>
    select(
      Core                         = core_id,
      `Timeframe (yr)`             = timeframe_yr,
      `Max age (yr BP)`            = maxage,
      `Rate — timeframe (g C/m²/yr)`  = seq_rate_g_m2_yr,
      `Rate — whole core (g C/m²/yr)` = seq_rate_wc_g_m2_yr
    ) |>
    kable(caption = "Estimated OC sequestration rates per core.")
}
```

### Stratum summary

```{r eval=has_rates}
if (has_rates && !is.null(sr_summary)) {
  sr_summary |>
    mutate(across(where(is.numeric), \(x) round(x, 1))) |>
    select(
      Stratum                           = stratum,
      `N cores`                         = n_cores,
      `Mean rate (g C/m²/yr)`           = mean_seq_rate_g_m2_yr,
      `SD (g C/m²/yr)`                  = sd_seq_rate_g_m2_yr,
      `Mean max age (yr BP)`            = mean_maxage_yr
    ) |>
    kable(caption = "Per-stratum mean OC sequestration rate.")
}
```

### SOC profiles with age annotations

The plot below shows the SOC concentration profile for each chronology core,
with sample ages labelled to illustrate how much carbon was accumulated over
the dated interval.

```{r eval=has_rates, fig.height=5, fig.cap="SOC concentration vs decompacted depth, annotated with interpolated sample ages."}
if (has_rates) {
  sr_cores |>
    filter(!is.na(age)) |>
    mutate(soc_g_kg = soc_pct * 10) |>
    ggplot(aes(x = soc_g_kg, y = depth_mid)) +
    geom_path(colour = "#2c7a4b", linewidth = 0.8) +
    geom_point(aes(colour = age), size = 2.5) +
    scale_y_reverse(name = "Decompacted depth (cm)") +
    scale_x_continuous(name = "SOC (g/kg)") +
    scale_colour_viridis_c(name = "Age (yr BP)", option = "plasma") +
    facet_wrap(~core_id, scales = "free_x") +
    labs(title = "SOC profile with age annotations") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
}
```

---

## Interpretation notes

| Rate context | Typical range (g C/m²/yr) |
|---|---|
| Global tidal marshes | 100–250 |
| Global seagrass meadows | 50–190 |
| Global mangroves | 160–250 |

*Source: Ouyang & Lee (2014), Duarte et al. (2013)*

These rates reflect **historical accumulation** over the dated interval and
may differ from modern sequestration due to land-use change, sea-level rise,
or vegetation shifts. Whole-core rates (open circles in the plot above) are
particularly sensitive to the age of the deepest horizon.

---

*Rendered: `r format(Sys.time(), "%Y-%m-%d %H:%M")` |
BlueCarbon package: `r packageVersion("BlueCarbon")`*
