---
title: "Carbon Sequestration Rates"
subtitle: "Radiometric age-depth chronology and OC accumulation rates"
date: today
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
if (!file.exists("_targets_seqrates")) setwd("..")
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

store <- "_targets_seqrates"

sr_cores      <- tar_read(sr_cores_with_ages,       store = store)
sr_rates      <- tar_read(sr_rates,                 store = store)
sr_summary    <- tar_read(sr_summary,               store = store)
sr_plot       <- tar_read(sr_plot,                  store = store)
sr_pb210_ages <- tar_read(sr_pb210_ages,            store = store)
sr_age_plot   <- tar_read(sr_pb210_age_plot,        store = store)
sr_crs_rates  <- tar_read(sr_crs_rates,             store = store)
sr_cic_rates  <- tar_read(sr_cic_rates,             store = store)
sr_comparison <- tar_read(sr_method_comparison,     store = store)
sr_comp_plot  <- tar_read(sr_method_comparison_plot, store = store)

has_rates <- !is.null(sr_rates) && nrow(sr_rates) > 0
has_pb210 <- !is.null(sr_pb210_ages) && nrow(sr_pb210_ages) > 0
```

## What this report shows

Carbon sequestration rate (CAR) is the mean rate at which organic carbon has
accumulated in the soil column over a defined time window, estimated from
radiometric ^210^Pb chronology data.

This report compares **three age-depth models** and shows how the choice of
model affects the estimated sequestration rate:

| Model | Input data | Assumption | Uncertainty |
|---|---|---|---|
| **Linear interpolation** | Pre-assigned dates (`core_chronology.csv`) | Constant rate between anchor points | None propagated |
| **CRS** (Constant Rate of Supply) | Raw ^210^Pb activity (`core_pb210.csv`) | Constant Pb flux; variable sedimentation | Full propagation |
| **CIC** (Constant Initial Concentration) | Raw ^210^Pb activity (`core_pb210.csv`) | Constant sedimentation; variable Pb flux | Full propagation |

**Common workflow:**

1. Sample depths are decompaction-corrected (same as the BlueCarbon pipeline).
2. Each age model assigns an age to every sample depth.
3. BlueCarbon's `estimate_seq_rate()` integrates OC accumulation over the
   timeframe (default **`r if(has_rates) sr_rates$timeframe_yr[1] else 100` years**).

---

## Chronology overview

### Cores with chronology data

```{r}
chrono_summary <- sr_cores |>
  filter(!is.na(age)) |>
  group_by(core_id) |>
  summarise(
    n_samples     = n(),
    max_depth_cm  = round(max(depth_mid, na.rm = TRUE), 1),
    max_age_yr    = round(max(age,       na.rm = TRUE), 0),
    mean_acc_rate = round(max(age, na.rm=TRUE) / max(depth_mid, na.rm=TRUE), 1),
    .groups = "drop"
  )

chrono_summary |>
  select(
    Core                             = core_id,
    `Samples with ages`              = n_samples,
    `Max depth (cm)`                 = max_depth_cm,
    `Max age (yr BP)`                = max_age_yr,
    `Mean accumulation (yr/cm)`      = mean_acc_rate
  ) |>
  kable(caption = "Summary of cores with radiometric chronology data.")
```

### Age-depth relationships

```{r fig.height=4, fig.cap="Age-depth profile for each chronology core. Lines show the interpolated age model used to assign ages to samples."}
sr_cores |>
  filter(!is.na(age)) |>
  ggplot(aes(x = age, y = depth_mid, colour = core_id, group = core_id)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_y_reverse(name = "Depth (cm)") +
  scale_x_continuous(name = "Age (years BP)") +
  scale_colour_brewer(palette = "Set2", name = "Core") +
  labs(title = "Age-depth model (linear interpolation)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

---

## Sequestration rates

```{r results="asis"}
if (!has_rates) {
  cat("> **No sequestration rates could be calculated.**  \n",
      "> This occurs when no core spans the full timeframe ",
      if (!is.null(sr_rates)) sr_rates$timeframe_yr[1] else 100,
      " years.  \n",
      "> Add deeper cores or increase age-depth coverage to enable rate estimation.\n")
}
```

### Per-core rates

```{r eval=has_rates, fig.cap="Carbon sequestration rates per core. Bar = rate over the target timeframe; circle = whole-core rate (total OC / max core age)."}
if (has_rates) sr_plot
```

```{r eval=has_rates}
if (has_rates) {
  sr_rates |>
    mutate(across(where(is.numeric), \(x) round(x, 1))) |>
    select(
      Core                         = core_id,
      `Timeframe (yr)`             = timeframe_yr,
      `Max age (yr BP)`            = maxage,
      `Rate — timeframe (g C/m²/yr)`  = seq_rate_g_m2_yr,
      `Rate — whole core (g C/m²/yr)` = seq_rate_wc_g_m2_yr
    ) |>
    kable(caption = "Estimated OC sequestration rates per core.")
}
```

### Stratum summary

```{r eval=has_rates}
if (has_rates && !is.null(sr_summary)) {
  sr_summary |>
    mutate(across(where(is.numeric), \(x) round(x, 1))) |>
    select(
      Stratum                           = stratum,
      `N cores`                         = n_cores,
      `Mean rate (g C/m²/yr)`           = mean_seq_rate_g_m2_yr,
      `SD (g C/m²/yr)`                  = sd_seq_rate_g_m2_yr,
      `Mean max age (yr BP)`            = mean_maxage_yr
    ) |>
    kable(caption = "Per-stratum mean OC sequestration rate.")
}
```

### SOC profiles with age annotations

The plot below shows the SOC concentration profile for each chronology core,
with sample ages labelled to illustrate how much carbon was accumulated over
the dated interval.

```{r eval=has_rates, fig.height=5, fig.cap="SOC concentration vs decompacted depth, annotated with interpolated sample ages."}
if (has_rates) {
  sr_cores |>
    filter(!is.na(age)) |>
    mutate(soc_g_kg = soc_pct * 10) |>
    ggplot(aes(x = soc_g_kg, y = depth_mid)) +
    geom_path(colour = "#2c7a4b", linewidth = 0.8) +
    geom_point(aes(colour = age), size = 2.5) +
    scale_y_reverse(name = "Decompacted depth (cm)") +
    scale_x_continuous(name = "SOC (g/kg)") +
    scale_colour_viridis_c(name = "Age (yr BP)", option = "plasma") +
    facet_wrap(~core_id, scales = "free_x") +
    labs(title = "SOC profile with age annotations") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
}
```

---

## ^210^Pb age model comparison

### Raw ^210^Pb activity profiles

The plot below shows the measured total ^210^Pb activity (Bq/kg) at each
depth. The supported (background) level of `r if(has_pb210) round(mean(tar_read(sr_pb210_data, store=store)$pb210_supported_Bq_kg),1) else "~15"` Bq/kg is indicated
by the dashed line. Excess activity above this level is used by both the
CRS and CIC models to derive ages.

```{r eval=has_pb210, fig.height=4, fig.cap="Raw total 210Pb activity profiles. Dashed line = supported (background) level. Activity declines with depth as the unsupported 210Pb decays."}
if (has_pb210) {
  pb210_data <- tar_read(sr_pb210_data, store = store)
  supported  <- mean(pb210_data$pb210_supported_Bq_kg)

  ggplot(pb210_data,
    aes(x = pb210_total_Bq_kg, y = depth_mid_cm, colour = core_id)) +
    geom_vline(xintercept = supported, linetype = "dashed", colour = "grey50") +
    geom_errorbarh(
      aes(xmin = pb210_total_Bq_kg - pb210_total_sd_Bq_kg,
          xmax = pb210_total_Bq_kg + pb210_total_sd_Bq_kg),
      height = 0.8, alpha = 0.5
    ) +
    geom_point(size = 2.5) +
    scale_y_reverse(name = "Depth (cm)") +
    scale_x_continuous(name = "Total ²¹⁰Pb activity (Bq/kg)") +
    scale_colour_brewer(palette = "Set2", name = "Core") +
    facet_wrap(~core_id) +
    labs(title = "Total ²¹⁰Pb activity with depth") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
}
```

### Age-depth models

```{r eval=has_pb210, fig.height=5, fig.cap="Age-depth comparison: CRS (green, with ±1 SD ribbon), CIC (orange), and linear interpolation (blue). CRS allows sedimentation rate to vary; CIC assumes it is constant."}
if (has_pb210) sr_age_plot
```

#### Model characteristics

**CRS (Constant Rate of Supply — Appleby & Oldfield 1978)**

Ages are derived from the cumulative ^210^Pb inventory remaining below each
horizon: older horizons have a larger fraction of the total inventory below
them. This model makes no assumption about sedimentation rate and handles
changes in accumulation well. It is the most widely used model in blue
carbon research.

**CIC (Constant Initial Concentration — Krishnaswami et al. 1971)**

Ages are derived from the log-linear decay of excess ^210^Pb with depth,
assuming a constant sedimentation rate. If sedimentation has varied (e.g.
post-disturbance acceleration), CIC will misdate those sections. Useful as
a check: large CRS–CIC age differences indicate non-constant accumulation.

**Linear interpolation (current default)**

Anchor ages from the `core_chronology.csv` file are linearly interpolated.
These anchors may themselves come from CRS modelling done externally, or
from ^14^C dates at depth. No measurement uncertainty is propagated.

```{r eval=has_pb210}
if (has_pb210) {
  sr_pb210_ages |>
    mutate(across(where(is.numeric), \(x) round(x, 1))) |>
    select(
      Core              = core_id,
      `Depth mid (cm)`  = depth_mid_cm,
      `CRS age (yr BP)` = age_crs,
      `CRS ±1 SD`       = age_crs_sd,
      `CIC age (yr BP)` = age_cic,
      `CIC ±1 SD`       = age_cic_sd
    ) |>
    kable(caption = "Model-derived ages at each ²¹⁰Pb measurement depth.")
}
```

---

## Sequestration rate comparison

The plot below shows the sequestration rate estimated from each age model
for the same cores and the same OC data. Differences between methods reflect
how much the age-depth model choice influences the final rate estimate.

```{r eval=has_rates && has_pb210, fig.height=4, fig.cap="Sequestration rate by core and age model. Bar = timeframe rate (100 yr); circle = whole-core rate."}
if (has_rates && has_pb210) sr_comp_plot
```

```{r eval=has_rates && has_pb210}
if (has_rates && has_pb210) {
  sr_comparison |>
    mutate(across(where(is.numeric), \(x) round(x, 1))) |>
    select(
      `Age model`                      = method,
      Core                             = core_id,
      `Rate — timeframe (g C/m²/yr)`   = seq_rate_g_m2_yr,
      `Rate — whole core (g C/m²/yr)`  = seq_rate_wc_g_m2_yr
    ) |>
    arrange(Core, `Age model`) |>
    kable(caption = "Sequestration rate by age model and core.")
}
```

### Which model to use?

| Situation | Recommended model |
|---|---|
| Variable sedimentation (disturbance, sea-level change) | **CRS** |
| Stable, uniform sediment with constant accumulation | CIC (as check against CRS) |
| ^14^C dates available at depth | Linear interpolation of calibrated ages |
| No raw activity data, only anchor dates | Linear interpolation (current default) |

For reporting to carbon registries (e.g. Verra VM0033), the CRS model is
generally preferred because it does not require the constant-sedimentation
assumption that is often violated in dynamic coastal wetlands.

---

## Interpretation notes

| Rate context | Typical range (g C/m²/yr) |
|---|---|
| Global tidal marshes | 100–250 |
| Global seagrass meadows | 50–190 |
| Global mangroves | 160–250 |

*Source: Ouyang & Lee (2014), Duarte et al. (2013)*

These rates reflect **historical accumulation** over the dated interval and
may differ from modern sequestration due to land-use change, sea-level rise,
or vegetation shifts. Whole-core rates (open circles in the plot above) are
particularly sensitive to the age of the deepest horizon.

---

*Rendered: `r format(Sys.time(), "%Y-%m-%d %H:%M")` |
BlueCarbon package: `r packageVersion("BlueCarbon")`*
