---
title: "Step 1: Non-Spatial Analysis"
subtitle: "Blue Carbon MMRV Workflow"
date: today
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# Load all packages required by this report.
# These must be declared here even though they are also in tar_option_set()
# in _targets.R — the knitr session is separate from the targets session.
library(targets)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)

# Pull all Step 1 targets from the _targets/ cache.
# tar_read() reads the cached object — it does NOT re-run analysis.
# Run tar_make() before rendering if any targets are outdated.
cfg              <- tar_read(cfg)
cores_raw        <- tar_read(cores_raw)
cores_clean      <- tar_read(cores_clean)
eda_plots        <- tar_read(eda_plots)
cores_harmonized <- tar_read(cores_harmonized)
stratum_summary  <- tar_read(stratum_summary)
```

## Project overview

```{r}
cores_qa <- cores_clean |> filter(qa_pass)
tibble(
  Metric = c("Project name", "Scenario", "Monitoring year",
             "Total cores (QA-passed)", "Total samples (QA-passed)",
             "QA pass rate (%)"),
  Value  = c(cfg$PROJECT_NAME, cfg$PROJECT_SCENARIO,
             as.character(cfg$MONITORING_YEAR),
             as.character(n_distinct(cores_qa$core_id)),
             as.character(nrow(cores_qa)),
             as.character(round(100 * mean(cores_clean$qa_pass, na.rm = TRUE), 1)))
) |> kable(col.names = c("", ""), align = c("l", "r"))
```

## QA/QC

### Summary

```{r}
tibble(
  Flag = c("Total samples", "Passed QA",
           "Failed: depth invalid", "Failed: SOC out of range",
           "Failed: BD out of range", "Failed: missing location"),
  `n samples` = c(
    nrow(cores_clean), sum(cores_clean$qa_pass),
    sum(!cores_clean$qa_depth_valid,  na.rm = TRUE),
    sum(!cores_clean$qa_soc_valid,    na.rm = TRUE),
    sum(!cores_clean$qa_bd_valid,     na.rm = TRUE),
    sum(!cores_clean$qa_has_location, na.rm = TRUE)
  )
) |> kable()
```

```{r fig.cap="Samples flagged per QA issue type. A sample can be flagged for multiple issues."}
eda_plots$qa_summary
```

### Bulk density source

```{r}
cores_clean |>
  filter(qa_pass) |>
  group_by(stratum) |>
  summarise(
    `n samples`    = n(),
    `BD measured`  = sum(!bd_estimated, na.rm = TRUE),
    `BD estimated` = sum( bd_estimated, na.rm = TRUE),
    `% estimated`  = round(100 * mean(bd_estimated, na.rm = TRUE), 1),
    `Mean BD`      = round(mean(bulk_density_g_cm3, na.rm = TRUE), 3),
    .groups = "drop"
  ) |>
  kable(
    col.names = c("Stratum", "n samples", "BD measured", "BD estimated",
                  "% estimated", "Mean BD (g/cm³)"),
    caption = paste(
      "BD defaults from blue_carbon_config.R BD_DEFAULTS:",
      paste(names(cfg$BD_DEFAULTS), unlist(cfg$BD_DEFAULTS),
            sep = " = ", collapse = ", ")
    )
  )
```

## Core locations

```{r fig.cap="Core locations coloured by stratum."}
eda_plots$spatial_map
```

## SOC profiles

```{r fig.cap="Raw SOC depth profiles. Each line is one core. QA-passed only."}
eda_plots$depth_profiles
```

```{r fig.cap="SOC distribution by stratum."}
eda_plots$soc_by_stratum
```

## Bulk density

```{r fig.cap="BD distributions. Dashed = literature defaults (bd_estimated = TRUE)."}
eda_plots$bd_distribution
```

## Carbon stocks (raw, pre-harmonization)

```{r fig.cap="Total raw carbon stock per core, summed across measured depth intervals."}
eda_plots$carbon_stocks
```

## Depth harmonization

### Method

Profiles harmonized to VM0033 standard depths using a hybrid approach:

- **Interpolation** (within measured range): monotone Hermite spline (`monoH.FC`).
- **Extrapolation** (beyond measured range): exponential decay if Spearman ρ < −0.3
  and n ≥ 3 per core; constant otherwise. Capped at 2.5× maximum measured depth.

Target depth midpoints: `r paste(cfg$VM0033_DEPTH_MIDPOINTS, "cm", collapse = ", ")`.
QC thresholds: SOC `r cfg$QC_SOC_MIN`–`r cfg$QC_SOC_MAX` g/kg;
BD `r cfg$QC_BD_MIN`–`r cfg$QC_BD_MAX` g/cm³.

### Results

```{r}
stratum_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  select(
    Stratum              = stratum,
    `Depth (cm)`         = depth_cm_midpoint,
    `n cores`            = n_cores,
    `Mean stock (kg/m²)` = mean_stock,
    `SD (kg/m²)`         = sd_stock,
    `Mean SOC (g/kg)`    = mean_soc,
    `Mean BD (g/cm³)`    = mean_bd,
    `% extrapolated`     = pct_extrapolated
  ) |>
  kable(caption = "Harmonized carbon stocks at VM0033 standard depth intervals.")
```

### Spline fit diagnostics

```{r}
cores_harmonized |>
  distinct(core_id, stratum, rmse, r2, qa_monotonic) |>
  group_by(stratum) |>
  summarise(
    `n cores`             = n(),
    `Mean RMSE`           = round(mean(rmse, na.rm = TRUE), 3),
    `Mean R²`             = round(mean(r2,   na.rm = TRUE), 3),
    `Non-monotonic cores` = sum(!qa_monotonic, na.rm = TRUE),
    .groups = "drop"
  ) |>
  kable(caption = "Non-monotonic = SOC Spearman rho >= -0.3 (not decreasing with depth).")
```

---

*Rendered: `r format(Sys.time(), "%Y-%m-%d %H:%M")` |
Pipeline: `r nrow(tar_meta())` targets*
