---
title: "Blue Carbon Analysis — Steps 1, 2 & 3"
subtitle: "Non-Spatial Analysis, Simple Extrapolation, and Random Forest Mapping"
date: today
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# Quarto renders from the document's directory; move up to project root
# so tar_read() can find the _targets/ store.
if (!file.exists("_targets")) setwd("..")

library(targets)
library(dplyr)
library(ggplot2)
library(knitr)

cfg                 <- tar_read(cfg)
cores_raw           <- tar_read(cores_raw)
eda_plots           <- tar_read(eda_plots)
cores_harmonized    <- tar_read(cores_harmonized)
stratum_summary     <- tar_read(stratum_summary)
step2_extrapolation <- tar_read(step2_extrapolation)
rf_models           <- tar_read(rf_models)
rf_importance_plot  <- tar_read(rf_importance_plot)
rf_maps             <- tar_read(rf_maps)
```

## Project overview

```{r}
tibble(
  Metric = c("Project name", "Scenario", "Monitoring year",
             "Total cores", "Total samples"),
  Value  = c(cfg$PROJECT_NAME, cfg$PROJECT_SCENARIO,
             as.character(cfg$MONITORING_YEAR),
             as.character(n_distinct(cores_raw$core_id)),
             as.character(nrow(cores_raw)))
) |> kable(col.names = c("", ""), align = c("l", "r"))
```

## Core locations

```{r fig.cap="Core locations coloured by stratum."}
eda_plots$spatial_map
```

## SOC profiles

```{r fig.cap="Raw SOC depth profiles. Each line is one core."}
eda_plots$depth_profiles
```

## Carbon stocks (raw)

```{r fig.cap="Total raw carbon stock per core, summed across measured depth intervals."}
eda_plots$carbon_stocks
```

## Depth harmonization

### Method

Profiles harmonized to VM0033 standard depths using a hybrid approach:

- **Interpolation** (within measured range): monotone Hermite spline (`monoH.FC`).
- **Extrapolation** (beyond measured range): exponential decay if Spearman ρ < −0.3
  and n ≥ 3 per core; constant otherwise. Capped at 2.5× maximum measured depth.

Target depth midpoints: `r paste(cfg$VM0033_DEPTH_MIDPOINTS, "cm", collapse = ", ")`.

### Results

```{r}
stratum_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  select(
    Stratum              = stratum,
    `Depth (cm)`         = depth_cm_midpoint,
    `n cores`            = n_cores,
    `Mean stock (kg/m²)` = mean_stock,
    `SD (kg/m²)`         = sd_stock,
    `Mean SOC (g/kg)`    = mean_soc,
    `Mean BD (g/cm³)`    = mean_bd,
    `% extrapolated`     = pct_extrapolated
  ) |>
  kable(caption = "Harmonized carbon stocks at VM0033 standard depth intervals.")
```

## Step 2 — Simple extrapolation

Mean harmonized stock per stratum summed across VM0033 depth intervals.
Total stock columns appear once `AOI_FILE` is set in `blue_carbon_config.R`.

```{r}
step2_extrapolation |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  kable(caption = "Per-stratum carbon stock summary.")
```

## Step 3 — Random Forest spatial prediction

### Model performance (80/20 holdout)

```{r}
rf_models$cv_metrics |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable(col.names = c("Outcome", "n test cores", "RMSE (kg/m²)", "R²"),
        caption = "Random Forest cross-validation on held-out 20% of cores.")
```

### Variable importance

```{r fig.cap="Top predictors for total carbon stock. Higher % increase in MSE = more important."}
rf_importance_plot
```

### Predicted carbon stock — by depth

```{r fig.cap="Random Forest predictions at each VM0033 depth interval (kg C/m²).", fig.width=10, fig.height=8}
rf_maps$depth_panel
```

### Predicted carbon stock — total to 1 m

```{r fig.cap="Total predicted carbon stock summed across all VM0033 depth intervals (kg C/m²)."}
rf_maps$total
```

---

*Rendered: `r format(Sys.time(), "%Y-%m-%d %H:%M")` |
Pipeline: `r nrow(tar_meta())` targets*
