---
title: "Blue Carbon Analysis — Step 4"
subtitle: "Transfer Learning with Global Coastal Wetland Data"
date: today
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# Look for the transfer store relative to project root
if (!file.exists("_targets_transfer")) setwd("..")
library(targets); library(dplyr); library(ggplot2); library(knitr)

cfg_tl    <- tar_read(cfg_tl,    store = "_targets_transfer")
tl_models <- tar_read(tl_models, store = "_targets_transfer")
tl_maps   <- tar_read(tl_maps,   store = "_targets_transfer")
```

## What this step does

Step 4 uses **transfer learning** to improve carbon stock predictions in a
local estuary with very few field cores (typically N ≈ 6–15) by borrowing
information from a global coastal wetland database.

**The challenge:** a random forest trained on 6 local cores can describe those
cores but cannot reliably generalise to unsampled pixels. The global database
has thousands of coastal wetland cores, but it was collected across many
different ecosystems and climates.

**The solution — Wadoux instance weighting:**

1. Train a *domain classifier* to identify which global cores are most similar
   to the local environment (using remote sensing covariates).
2. Up-weight similar global cores so they dominate the global random forest.
3. Apply the global random forest to the local site as a *prior*.
4. Estimate a *bias correction* — the average difference between what the
   global model predicts and what was actually measured locally.
5. Add the bias correction to every predicted pixel.
6. Bootstrap the local residuals to generate 90% prediction intervals
   (required by VM0033).

**Outputs:**

- Per-pixel maps for each VM0033 depth interval showing the global prior and
  the bias-corrected (transfer) prediction.
- A LOCO (leave-one-core-out) cross-validation table comparing transfer
  learning against the uncorrected global model.

## Model performance

```{r cv-table}
summary_df <- tl_maps$summary

if (nrow(summary_df) == 0) {
  cat("No models were successfully trained. Check data completeness.")
} else {
  summary_df |>
    mutate(
      depth_interval = c("0–15 cm", "15–30 cm", "30–50 cm", "50–100 cm")[
        match(depth_cm, c(7.5, 22.5, 40, 75))
      ],
      bias_correction = sprintf("%+.3f", bias_correction),
      cv_r2_tl        = ifelse(is.na(cv_r2_tl),     "—", sprintf("%.3f", cv_r2_tl)),
      cv_r2_global    = ifelse(is.na(cv_r2_global),  "—", sprintf("%.3f", cv_r2_global)),
      cv_rmse_tl      = ifelse(is.na(cv_rmse_tl),    "—", sprintf("%.4f", cv_rmse_tl)),
      cv_rmse_global  = ifelse(is.na(cv_rmse_global),"—", sprintf("%.4f", cv_rmse_global))
    ) |>
    select(
      `Depth interval` = depth_interval,
      `N global`   = n_global,
      `N local`    = n_local,
      `Covariates` = n_covariates,
      `Bias (kg/m²)` = bias_correction,
      `Bias SE`    = bias_se,
      `TL R²`      = cv_r2_tl,
      `Global R²`  = cv_r2_global,
      `TL RMSE`    = cv_rmse_tl,
      `Global RMSE`= cv_rmse_global
    ) |>
    kable(align = c("l", rep("r", 9)),
          caption = "LOCO CV: transfer learning vs uncorrected global model")
}
```

### Interpreting the table

| Column | What it means |
|--------|---------------|
| **Bias** | Mean residual: how much the global model over/under-predicts locally. Added to all pixels in the Transfer Final map. |
| **Bias SE** | Bootstrap standard error on the bias estimate (500 resamples). Larger values indicate the local sample is too small to pin down the bias reliably. |
| **TL R²** | Cross-validation R² after bias correction. Negative values are possible when N_local is tiny and the global model is biased. |
| **Global R²** | Same but without the bias correction. A positive improvement from Global to TL confirms the correction is helping. |

> **VM0033 guidance:** the standard requires uncertainty at the 90% confidence
> level. Total prediction SE = √(bias_SE² + residual_variance). With N_local < 10,
> the bias SE dominates — collect more cores to narrow this interval.

## Maps

### Global prior vs transfer-corrected prediction

```{r maps, fig.height=8}
print(tl_maps$maps)
```

**Global prior** — the Wadoux-weighted random forest trained on the global
database, with no local correction.

**Transfer (bias-corrected)** — global prior shifted by the mean local
residual. This map accounts for systematic over- or under-prediction in the
local ecosystem.

### Cross-validation comparison

```{r cv-plot}
print(tl_maps$validation)
```

## How to use these outputs

**For VM0033 reporting:**

The Transfer Final rasters (`outputs/transfer/tl_carbon_stocks_kg_m2.tif`)
are the recommended prediction surfaces. Bands per depth are named:

- `dX_Global_Prior` — uncorrected global RF prediction
- `dX_Transfer_Final` — bias-corrected prediction **(use this)**
- `dX_Local_Only` — constant raster = mean of local cores (naive baseline)
- `dX_Difference` — Transfer Final minus Local Only

To generate 90% prediction intervals for a depth d:

```r
m   <- tl_models$models[["7.5"]]
se  <- sqrt(m$bias_se^2 + m$residual_sd^2)
z   <- qnorm(0.95)   # one-sided 95% = two-sided 90% CI

# For a single pixel with transfer_final prediction p:
lower <- p - z * se
upper <- p + z * se
```

**Next steps:**

- Run Step 4 → Step 3 comparison: do the transfer learning maps and the local
  RF agree on the spatial pattern?
- If bias is large (|bias| > 0.5 × local mean), investigate whether the global
  training data covers the local ecosystem type adequately.
- Add more local cores to reduce bias SE and narrow the 90% CI.
